Instructions for developing new pages for the MarineOmics webpage

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

---
title: 'DNA Methylation Assessment'
subtitle: '*Yaamini Venkataraman*, *Sam White*, and *Steven Roberts*'
bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
---

@@

    -   **`--multicore 28`**: Specifies that the program should use 28 cores to process the data in parallel, which can significantly speed up the analysis.

    -   **`--buffer_size 75%`**: This option sets the buffer size for reading in BAM files to 75% of the available system memory. It helps optimize the performance of the tool by adjusting how much memory is used for processing.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format). The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined. Later in this workflow, a bedGraph with *merged* information will be created.

@@

A genome-wide cytosine report is generated (including merged bedGraph file), with this code as an example:

@@

@@

From here it could be useful to do some simple file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg bedtools, GLMs).

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
  > "${STEM}"_10x.bedgraph
done

for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
  > "${STEM}"_5x.bedgraph
done
```

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_10x.tab
done

for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_5x.tab
done
```

@@

# Next Steps & Downstream applications

## CT SNPs

Masking cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to considerfor a few reasons:

1.  **Bisulfite Conversion Process**: In bisulfite sequencing, unmethylated cytosines are converted to uracil, which is then read as thymine in sequencing. This conversion is key to identifying methylated cytosines. However, if a genomic DNA sequence naturally has a C to T SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.

2.  **Reduction of False Positives**: By masking C to T SNPs, one can reduce false positives in methylation analysis. Without masking, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.

3.  **Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs). Masking C to T SNPs allows for a more accurate representation of the methylation status of the genome.

4.  **Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, masking C to T SNPs ensures consistency across samples. This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.

5.  **Enhanced Data Interpretation**: By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.

One way to do this is using BS-Snper (@10.1093/bioinformatics/btv507). <https://github.com/hellbelly/BS-Snper>. In short one could merge all BAM files from experiment run BS-Snper, filter out CT SNPS then use this to remove loci.

@@

10. **`--minread2 2`**: The minimum number of reads required to support a SNP candidate.

>>>>

---
title: 'DNA Methylation Assessment'
subtitle: '*Yaamini Venkataraman*, *Kathryn Greiner-Ferris*, *Sam White*, and *Steven Roberts*'
bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
---

@@

    -   **`--multicore 28`**: Specifies that the program should use 28 cores to process the data in parallel, which can significantly speed up the analysis. This can be adjusted depending on your computing power.

    -   **`--buffer_size 75%`**: This option sets the buffer size for reading in BAM files to 75% of the available system memory. It helps optimize the performance of the tool by adjusting how much memory is used for processing. This can be adjusted depending on your computing power.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format):

@@

The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined. It is important to merge the reads from the forward and reserve strands for more accurate methylation quantification. A genome-wide cytosine report and merged bedGraph file are generated (including merged bedGraph file) from the `deduplicated.bismark.cov.gz` generated above in this example code:

@@

# Summary report

@@

From here it could be useful to do some file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg `bedtools`, GLMs). Specifically, it may be important to filter the CpG methylation data to a specific coverage threshold (ex. 5x or 10x). The more read coverage a CpG has, the higher the confidence in methylation quantification.

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
  > "${STEM}"_5x.bedgraph
done
```

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_5x.tab
done
```

@@

# Next Steps

So you've aligned your trimmed reads to your reference genome and have methylation data for CpGs. Now what?

## Removing C to T (C->T) SNPs

Removing cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to consider for a few reasons:

1.  **Bisulfite Conversion Process**: In bisulfite sequencing, unmethylated cytosines are converted to uracil, which is then read as thymine in sequencing. This conversion is key to identifying methylated cytosines. However, if a genomic DNA sequence naturally has a C to T (C->T) SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.

2.  **Reduction of False Positives**: By removing C->T SNPs, one can reduce false positives in methylation analysis. Without removal, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.

3.  **Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs). Removing C->T SNPs allows for a more accurate representation of the methylation status of the genome.

4.  **Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, removing C->T SNPs ensures consistency across samples. This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.

By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.

One way to do this is using BS-Snper (@10.1093/bioinformatics/btv507). <https://github.com/hellbelly/BS-Snper>. In short one could merge all BAM files from experiment run BS-Snper, filter out C->T SNPS then use this to remove loci.

@@

10. **`--minread2 2`**: The minimum number of reads required to support a SNP candidate. It could be useful to use the same values for the minimum read threshold for SNP identification and for minimum coverage in the methylation data.
